---
title: "Reproducible Research: Peer Assessment 1"
author: "William Thong"
date: "21st Dec 2016"
output: 
  html_document:
    keep_md: true
---

```{r LoadLibs, echo=FALSE}

# Load libraries
library(dplyr)
library(ggplot2)
library(lubridate)

```


## Loading and preprocessing the data
```{r ProcessData}

# Ensure all files are in the intended working dir else use setwd() to set
getwd()
#setwd("your working directory")

# Check, download & unzip raw data file if required
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
if (!file.exists("./activity.zip")){
  download.file(fileUrl,"./activity.zip", mode="wb")
}
if (!file.exists("./activity.csv")){
  unzip("./activity.zip", exdir=".")
}

# Import data, convert date var to proper date format and str() to view
dfActivity <- read.csv("activity.csv", header=TRUE, stringsAsFactors=FALSE)
dfActivity$date <- ymd(dfActivity$date)
str(dfActivity)

# Run summary stats on imported data to get a view on the shape of the data
summary(dfActivity)

```


## What is mean total number of steps taken per day?  
1. Calculate the total number of steps taken per day.  
2. Make a histogram of the total number of steps taken each day.  
3. Calculate and report the mean and median of the total number of steps taken per day.  

```{r MeanTotalSteps}

# Calculate total number of steps taken per day using dplyr package
dfTotalSteps <- dfActivity %>%
        filter(!is.na(steps)) %>%
        group_by(date) %>%
        summarise(varTotalSteps = sum(steps))
str(dfTotalSteps)

# Calculate mean and median of steps
varMeanSteps <- mean(dfTotalSteps$varTotalSteps)
varMedianSteps <- median(dfTotalSteps$varTotalSteps)

# Plot histogram using ggplot2 package. Note: Changing the number of bins will result in different plots
plot.TotalSteps <- ggplot(dfTotalSteps, aes(varTotalSteps)) +
        geom_histogram(bins=20, fill=I("blue"), alpha=1/2, col=I("black")) +
        ggtitle("Frequency Distribution: Total Number of Steps Taken Per Day") +
        labs(x="Total Number of Steps Taken Per Day") +
        labs(y="Frequency Count")
print(plot.TotalSteps)

```
<br>
The histogram indicated that the distribution of total number of steps taken per day is right-skewed with the possibilties of extreme outliers.  The mean of the total number of steps taken per day is **`r paste(varMeanSteps)`** while the median is **`r paste(varMedianSteps)`**.  Because the distribution is skewed, the median should be used as the average.     

## What is the average daily activity pattern?  
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).  
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?  
```{r AvgPattern}

# Calculate the mean for the 5 mins intervals
df5MinIntervals <- dfActivity %>%
        filter(!is.na(steps)) %>%
        group_by(interval) %>%
        summarise(varAvgSteps = mean(steps))
str(df5MinIntervals)

# Find the 5-minute interval that conains the maximum number of steps
varMaxAvgSteps <- df5MinIntervals %>%
        filter(varAvgSteps == max(varAvgSteps))
str(varMaxAvgSteps)

# Plot time series graph
plot.AvgSteps <- ggplot(df5MinIntervals, aes(x=interval, y=varAvgSteps)) +
        geom_line(colour="blue") +
        ggtitle("Daily Average Steps by 5-minutes Intervals") +
        labs(x="5-minutes Intervals") +
        labs(y="Average Number of Steps") +
        geom_point(data=varMaxAvgSteps,color="red", size=3)
print(plot.AvgSteps)
        
```
<br>
The plot above is the **daily average steps by 5-minutes intervals** across all days.  The time interval **`r paste(varMaxAvgSteps$interval)`** minutes have the highest average number of steps at **`r paste(varMaxAvgSteps$varAvgSteps)`**.


## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
